A preliminary constraint grammar for Russian

### 2015-03-28 2015-nodalida-ruscg/ruscg.tex
learning curve for voting tagger
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59548 72bbbca6-d526-0410-a7d9-f06f51895060
måns' paper
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59549 72bbbca6-d526-0410-a7d9-f06f51895060
some intrinsic eval
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59551 72bbbca6-d526-0410-a7d9-f06f51895060

\usepackage{times}
\usepackage{latexsym}
\usepackage{fixltx2e} %allows subscripts
%\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\usepackage[small,bf]{caption}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

\subsection{Morphological analyser}
% ROB

@@

\ex. \label{ex:FSToutput} 
	\a. \rus{новый}+A+Msc+Inan+Sg+Nom
	\b. \rus{автомат}+N+Msc+Inan+Sg+Nom
	\c. 
% det/prn (e.g. pos... их, я, мой)
% 'over'generation (e.g. all dets get all gen/nbr/case combinations)

@@

\section{Evaluation}

\subsection{Corpus}

In order to evaluate the grammar we hand-annotated 10,150 words of Russian text
from Wikipedia articles, public domain literature and freely-available news sources. The
annotated texts are available online under the {\sc cc-by-sa} licence.\footnote{\url{http://...}}

@@

We chose our own texts as opposed to using a well-known hand-annotated corpus
such as the Russian National Corpus (RNC) for two main reasons: the first was that the 
RNC is not freely available; the second was that the standards for tokenisation, part-of-speech
and morphological description are different from our morphological analyser.

@@

\begin{table*}
  \centering
  \begin{tabular}{|l|r|r|r|r|r|}
    \hline
    \textbf{Domain} & \textbf{Tokens} & \textbf{Precision} & \textbf{Recall} & \textbf{F-score} & \textbf{Ambig. solved} \\
    \hline
    Wikipedia       & 7,857      & 0.506        & 0.996    & 0.671 & 44.92\%  \\
    Literature      & 1,652      & 0.473        & 0.984    & 0.638 & 42.95\%  \\
    News            & 642        & 0.471        & 0.990    & 0.638 & 41.60\%  \\
    \hline
    \textbf{Average}& 10,150     & 0.498        &  0.994   & 0.663 & 44.39\% \\
    \hline
  \end{tabular}
  \caption{Results for the test corpora.}
\end{table*}

@@

@@

@@

>>>>

\usepackage{times}
\usepackage{latexsym}
\usepackage{fixltx2e} %allows subscripts
%\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{graphicx}
\usepackage[small,bf]{caption}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

\subsection{Morphological analyser}
\label{sec:morph}

@@

\ex. \label{ex:FSToutput} 
	\a. \rus{новый}\texttt{{\small <adj><m><nn><sg><nom>}}
	\b. \rus{автомат}\texttt{{\small <n><m><nn><sg><nom>}}
	\c. 
% det/prn (e.g. pos... их, я, мой)
% 'over'generation (e.g. all dets get all gen/nbr/case combinations)

@@

\section{Evaluation}
\label{sec:eval}

\subsection{Corpus}
In order to evaluate the grammar we hand-annotated 10,150 words of Russian text
from Wikipedia articles, public domain literature and freely-available news sources. The
annotated texts are available online under the {\sc cc-by-sa} licence.\footnote{\url{http://...}}

@@

We chose to annotate our own texts as opposed to using a well-known hand-annotated corpus
such as the Russian National Corpus (RNC) for two main reasons: the first was that the 
RNC is not freely available; the second was that the standards for tokenisation, part-of-speech
and morphological description are different from our morphological analyser.

@@

\begin{table*}
  \centering
  \begin{tabular}{|l|r|r|r|r|r|}
    \hline
    \textbf{Domain} & \textbf{Tokens} & \textbf{Precision} & \textbf{Recall} & \textbf{F-score} & \textbf{Ambig. solved} \\
    \hline
    Wikipedia       & 7,857      & 0.506        & 0.996    & 0.671 & 44.92\%  \\
    Literature      & 1,652      & 0.473        & 0.984    & 0.638 & 42.95\%  \\
    News            & 642        & 0.471        & 0.990    & 0.638 & 41.60\%  \\
    \hline
    \textbf{Average}& 10,150     & 0.498        &  0.994   & 0.663 & 44.39\% \\
    \hline
  \end{tabular}
  \caption{Results for the test corpora.}
  \label{table:results}
\end{table*}

@@

\begin{description}
  \item[ Bad linguistics:] In some cases a rule did not take into account grammatical possibilities 
    in the language. e.g. Two simple rules such as 
    \begin{itemize}
      \item \texttt{REMOVE Det IF (0 Det OR Pron) (1C Ne) ;}
      \item \texttt{REMOVE Det IF (0 Det OR Pron) (1 Cm LINK 1 CC OR CS) ;}
    \end{itemize}
    did not take into account the possibility of having a postposed determiner as in
    \begin{itemize}
      \item \ldots \rus{а может быть и раньше, и факт этот не раз поражал меня} \ldots
      \item \ldots and maybe even earlier, and fact \textbf{this} not once surprised me \ldots
    \end{itemize}
    or a interposed parenthetical as in
    \begin{itemize}
      \item \rus{Но какие, однако же, два разные создания, точно обе с двух разных планет!}
      \item But what, \textbf{exactly} , two different creatures, just both from two different planets!
    \end{itemize}
  \item[ Bad rule:] In some cases a rule was simply incorrectly specified. For example, the 
    following rule was designed to solve the ambiguity between short-form neuter adjectives and 
    adverbs
   \begin{itemize}
     \item \texttt{REMOVE A + Short IF (-1C Fin OR Adv OR A) (0C Short OR Adv) ;}
   \end{itemize}
   However there is no reason why we should prefer an adverb over an adjective after an adverb, 
   \begin{itemize}
     \item \ldots \rus{потому что совсем неприятно проснуться в гробу под землею.}
     \item \ldots because [it is] really \textbf{unpleasant} to wake up in a coffin under the ground.
   \end{itemize}
  \item[ Incomplete barrier:] xxx 
   \begin{itemize} 
     \item \texttt{REMOVE NGDAIP - Acc - Prp - Loc IF (-1C* Pr/V OR Pr/Na BARRIER (*) - Adv - Comp - DetIndecl - ModAcc - ModPrp) ;}
   \end{itemize}
   here the rule removes the nominative reading of the adjective to leave the accusative reading because the preposition \rus{в} `in'
   is found preceeding. 
   \begin{itemize}
     \item \rus{В 1960-х электрифицированные высокоскоростные железные дороги появились в Японии и некоторых других странах.}
     \item In the 1960's \textbf{electrified} high-speed railways appeared in Japan and some other countries.
   \end{itemize} 
  \item[ Incomplete set:] In some cases the rule was a good generalisation, but made use of a set
    which was incomplete. For example:
   \begin{itemize}
     \item \texttt{REMOVE Dat IF (NOT 0 Prn/Sebe) (NOT 0 Anim OR Cog OR Ant) (NOT 0 Pron) (NOT 1* V/Dat) (NOT -1* V/Dat) (NOT -1* Prep/Dat) (NOT -1C A + Dat) ;}
   \end{itemize} 
   the set \texttt{V/Dat} does not contain the verb \rus{противопоставлять} `opposed to' which takes 
   a dative argument.
   \begin{itemize}
     \item \rus{В связи с этим ортодоксальности стали противопоставлять ересь.}
     \item In connection with this \textbf{orthodoxy} was opposed to heresy. 
   \end{itemize}
  \item[ Rule interaction:] The strong accusative rule below causes incorrect behaviour in the rule to remove transitivity readings
  \begin{itemize}
     \item \texttt{REMOVE TV - Pass IF (NOT 1* Acc) (NOT -1* Acc) ;}
     \item \texttt{REMOVE Acc IF (-1C Fin + IV) (NOT 0 AccAdv) ;}
  \end{itemize}
  Consider the following example where \rus{может} `can' is tagged as intransitive, the second rule fires removing 
  the accusative reading of \rus{его} `him', and thus given the lack of accusative reading, \rus{найти} `find'
  is disambiguated as intransitive instead of transitive.
  \begin{itemize}
    \item \rus{Она смотрит везде, но не может его найти.}
    \item She looks around, but she cannot \textbf{find him}.
  \end{itemize}
  \item[ Difficult linguistics:] Dealing with participles with arguments is challenging in the case that the arguments 
    of the participle share the same government as the main verb.
  \begin{itemize}
    \item \texttt{REMOVE IV IF (0 TV OR IV) (1C Acc) (NOT 1 AccAdv) ;}
  \end{itemize}
    Here \rus{Ваню и Машу} `Vanja and Maša' are the object of \rus{видит} `sees' and not \rus{играющих} `playing', although
    both verbs can take accusative object.
  \begin{itemize}
    \item \rus{Их мама внутри дома с кошкой, она смотрит в окно и видит играющих Ваню и Машу.}
    \item Their mother is inside the house with the cat, she looks through the window and sees Vanja and Maša \textbf{playing}.
  \end{itemize}
    This kind of error would ideally be resolved with semantic knowledge. 
\end{description}

@@

Given that just over half of all ambiguity remains after running our preliminary constraint
grammar and that for many applications unambiguous output is necessary, we decided to 
experiment with combining the constraint grammar with a statistical tagger to resolve remaining
ambiguity.

We follow the voting method described by \newcite{hulden12}. We used the freely available
\texttt{hunpos} part-of-speech tagger \cite{halacsy07}. We performed 10-fold
cross validation using our evaluation corpus, taking 10\% for testing and 90\% for training, and
experimented with three configurations:

\begin{itemize}
  \item \texttt{HMM}: the \texttt{hunpos} part-of-speech tagger with its default options
  \item \texttt{HMM+Morph}: as with \texttt{HMM} but incorporating the output of our 
    morphological analyser (see section~\ref{sec:morph}) as a full form lexicon.
  \item \texttt{HMM+Morph+CG}: we submitted the output from \texttt{HMM+Morph} and the 
    constraint grammar to a voting procedure, whereby if the constraint grammar left
    one valid reading, we chose that, otherwise if the constraint grammar left a word
    with more than one reading, we chose the result from the \texttt{HMM+Morph} tagger.
\end{itemize}

As can be seen from Figure~\ref{fig:curve}, incorporating the constraint grammar 
improves the performance of the HMM tagger, an improvement of nearly 5\% in accuracy,
similar to that reported by \newcite{hulden12} for the same amount of training data. Our 
constraint grammar also has a much lower precision as a result of the ambiguity remaining
in the output.

Note that the final accuracy is below the state of the art for Russian, however the state of
the art employs unavailable training corpora of over five million tokens.

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{graphics/learning-curve.pdf}
  \caption{Learning curve for three taggers, \texttt{hunpos} with no lexicon, 
    \texttt{hunpos} with a lexicon, and \texttt{hunpos} with a lexicon and the 
    Russian constraint grammar in a voting set up.}
  \label{fig:curve}
\end{figure}

@@

% proof-of-concept for russian, but russian has loads of tagged corpora, for other langs, e.g. sorbian or belarusian 
% a CG+hmm could be much faster than tagging hundreds of thousands of words of text.
